---
title: "class12: Intro to Genome Informatics Lab"
author: "Paige Contreras (A16850883)"
format: pdf
---

## Section 1: Proportion og. G/G in a population


Downloaded a CSV from Ensemble <https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39695100-40095101;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel>

Here we read this CSV file

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378 (4).csv")
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) * 100
```

Now let's look at a different population. I picked the GBR population.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")

```

Find proportion of G|G

```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) *100,2 )
```

This variant that is associated with childhood asthma is more frequent in the GBR populaiton than the MKL population.

Lets now dig into this further.

## Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about

~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378…) on ORMDL3
expression.


How many sampples do we have?

```{r}
expr <-  read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```

```{r}
nrow(expr)
```


```{r}
table(expr$geno)
```

```{r}
library(ggplot2)

```

lets make a boxplot

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot()
```


```{r}
boxplot2 <- boxplot(exp ~ geno, data = expr)
```

```{r}
boxplot2$stats[3, ]
```

then to see which genotype belongs to which median...

```{r}
boxplot2$names
```

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes. 

The sample size for the correxsponding median expression levels for each genotype is:
"A/A"    "A/G"    "G/G"
31.24847 25.06486 20.07363

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

here is the boxplot once again...

```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot()
```

From this boxplot, we can infer that the G allele causes a decrease in expression as we see that expression levels go down when present. This is proven because A/A has the highest level of expression but once the G allele is introduced it goes down, and once again with the G/G genotype.

