---
title: "class08: Breast Cancer Mini Project"
author: "Paige Contreras (A16850883)"
format: pdf
toc: true
---

## Background 

In today's class we will be employing all the R techniques for data analysis that we have learned thus far - including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy.

## DATA Import 

The data is in CSV format

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names=1)
```

wee peak at the data
```{r}
head(wisc.df, 3)
```

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum( wisc.df$diagnosis =="M" )
```

```{r}
table( wisc.df$diagnosis )

```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.df)) )
```

```{r}
colnames(wisc.df)
```

we need to remove the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale=TRUE` here as the data columns, features/dimensions are on very different scales in the original data set. 

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
```

```{r}
head(wisc.pr$x)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

0.4427 is captured by the first principal component (PC1).

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3 are required for at least 70% of the original variance. 

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC6 are required for 90% of the original variance.

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is not easy to understand, the data is all mixed and uncomprehensible. You cannot see what it is trying to point out.

###Scree Plot 

```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```



> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

These plots are similar, but the new plot for component 1 and 3 had a downward shift and the points seem to be tighter, so there is less variance.


> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

The feature concave.points_mean has a negative value of -0.2608538 which means that higher values will contribute to PC1 negatively. Yes, there are features with larger contributions 

```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```

#4 Hierarchical clustering

The goal of this section is to do hierarchical clustering of the original data to see if there is any obvious grouping into malignant and benign clusters.

In short, these results are not good!

First we will scale our `wisc.data` then calculate a distance matrix, then pass to `hclust()`:

```{r}
wisc.dist <- dist(scale(wisc.data))
  wisc.hclust <- hclust(wisc.dist)
plot(wisc.hclust)
```



```{r}
 wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```

## 5 Combining methods

The idea here is that I can take my new variables (the PCs) that are better descreiptors of the data-set than the original features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```


```{r}
table(diagnosis)
```


I can now run `table()` with both my clustering `grps` and the expert `diagnosis`

```{r}
table(grps, diagnosis)
```
Our cluster "1" has 179 "M" diagnosis
Our cluster "2" has 333 "B" diagnosis 

179 TP
24 FP
333 TN
33 FN

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity: TN/(TN+FP)
```{r}
333/(333+24)
```

## Prediction

We can use our PCA model for prediction of new un-seen cases:

WHERE WE LEFT OFF IN CLASS^^


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
pve <- pr.var / sum(pr.var)
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

These plots are similar, but the new plot for component 1 and 3 had a downward shift and the points seem to be tighter, so there is less variance.


> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=19, col="pink", lty=2)
```

At height 19 there are four clusters.

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

Ward.D2 gives me the best results because it separates the clusters well within the variance. 

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

```{r}
table(wisc.pr.hclust.clusters, diagnosis) 
```


> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```
```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
Wisc.pr.hclust.clusters is better because it clusters the tumors correct more. 

> Q16. Which of these new patients should we prioritize for follow up based on your results?

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```


I think we should follow up with the patients that are plotted red. I think this because the group is clustered closer together which shows higher similarity 


