---
title: "class6hw6"
author: Paige Contreras
format: pdf
---

df <- data.frame(a=1:10, b=seq(200,400,length=10),c=11:20,d=NA)
df$a <- (df$a - min(df$a)) / (max(df$a) - min(df$a))
df$b <- (df$b - min(df$a)) / (max(df$b) - min(df$b))
df$c <- (df$c - min(df$c)) / (max(df$c) - min(df$c))
df$d <- (df$d - min(df$d)) / (max(df$a) - min(df$d)) 

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b
plotb3(s1.b, sse=s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```


> Q. What are the inputs to the function

The names are the IDs of the protein structures

> Q. what the function does and how to use it

The function gives us the plot of the protein structures.

> Q. What is the output of the function

The output is the plot from the protein ID that we input. It will also give us a bfactor value of the protein. 

```{r}
library(bio3d)
cluster_proteins <- function(pdb_id, chain = "A") {
  pdb <- read.pdb(pdb_id)
  pdb_chain <- trim.pdb(pdb, chain = chain, elet = "CA")
b <- pdb_chain$atom$b

plotb3(b, sse = pdb_chain, typ = "l", ylab = "B-factor")

return(b)
}
```
```{r}
 hc <- hclust( dist( rbind(s1.b, s2.b, s3.b) ) )
plot(hc)
```

```{r}
cluster_proteins("1AKE")
```


