---
title: "class07"
author: "Paige Contrersas (A16850883)"
format: pdf
---

## Background 
Today we will begin our exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**.

Let's make up some input data for clustering where we know what the natural "clusters" are.

The function `rnorm()` can be useful here.
## K-means clustering

```{r}
hist( rnorm(5000) )
```


> Q. Generate 30 random numbers centered at +3 and another 30 centered at -3

```{r}
rnorm(30, mean = 3)
rnorm(30, mean = -3)
```
```{r}
tmp <- c(rnorm(30, mean = 3),
rnorm(30, mean =-3) )

cbind(x=tmp, y =rev(tmp))
```

```{r}
cbind(letters, rev(letters))
```

```{r}
x <- cbind(x = tmp, y=rev(tmp))
plot(x)
```


## K means clustering 
The main function in "base R" for K-means clustering is called `kmeans()`:

```{r}
km <- kmeans(x, centers = 2)
```


> Q. What component of the results object details the cluster sizes

```{r}
km$size
```

> Q. What component of the results object dtails the cluster centers?

> What component of the results object details the cluster membership vector (i.e. our main result of which points lie in which cluster?)

```{r}
km$cluster
```

> Q. Plot our clustering results with points colors by cluster and also add the cluster centers as new points colored blue?

```{r}
plot(x, col=km$cluster )
points(km$centers, col="blue", pch = 15)
```


> Q. Run `kmeans()` again and this time produce 4 clusters (and call your result object `k4`) and make a results figure like above?

```{r}
k4 <- kmeans(x, centers = 4)
plot(x, col=k4$cluster )
points(k4$centers, col="blue", pch = 15)
```

the meteric
```{r}
km$tot.withinss
k4$tot.withinss
```

```{r}
ans <- NULL
for(i in 1:30) {
ans <- c(ans, kmeans(x, centers = i)$tot.withinss)
}


ans
```
```{r}
plot(ans, typ = "o")
```

**Key-point:** K-means will impose a clustering structure on your data even if it is not there - it will always give you the answer you asked for even if that anwer is silly!

##Hierarchical Clustering 

The main function for Hierarchical Clustering is called `hclust()`
unlike `kmeans()` (which does all the work for you) you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function. 

```{r}
d <- dist(x)
hc <- hclust(d)
plot(hc)
```

To extract our cluster membership vector from a `hclust()` result object we have to "cut" our tree at a given height to yield separate "groups"/"branches".

```{r}
plot(hc)
abline(h=8, col="pink", lty=)
```

to do this we use the `cutree()` function on our `hclust()` object:
```{r}
grps <- cutree(hc, h=8)
grps
```

```{r}
table(grps, km$cluster)
```
## PCA of UK food data

Import the dataset of food consumption in the UK

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

One solution to set the row names is to do it by hand...

```{r}
rownames(x) <- x[,1]
```
To remove the first column I can use the minus index trick

```{r}
x <- x[,-1]
x
```

A better way to do this is to set the row names to the first column by arguing with `read.csv()`

```{r}
x <- read.csv(url, row.names = 1)
x
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

### Spotting major differences and trends

is difficult even in this wee 17d dataset...

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```


```{r}
library(pheatmap)
pheatmap( as.matrix(x) )

```


##PCA to the rescue

The main PCA function in "base R" is called `prcomp()`. This function wants the transpose of our food data as inout (i.e. the foods as columns and the countries as rows). 

```{r}
pca <- prcomp( t(x) )
```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```
To make one of main PCA result figures we turn to `pca$x` the scores along our new PCs. This is called "PC plot" or "score plot" or "Ordienation plot"...

```{r}
pca$x
```

```{r}
my_cols <- c("peachpuff3", "red", "lightblue", "rosybrown1")
```


```{r}
library(ggplot2)
ggplot(pca$x) +
aes(PC1, PC2,) +
geom_point(col=my_cols)
```


The second major result figure is called a "loading plot" of "variable contributions plot" or "weight plot"


```{r}
ggplot(pca$rotation) +
aes(PC1, rownames(pca$rotation)) +
geom_col()

```






